Exploring probability distributions for bivariate temporal granularities

### 2019-11-25 main.Rmd
spell check rough draft

Calendar-based graphics[@wang2019tsibble] are useful in visualizing patterns in the weekly and monthly structure well and are capable of checking the weekends or special days.  Any sub-daily resolution temporal data can also be displayed using this type of faceting [@Wickham2009-pk] with days of the week, month of the year and another sub-daily deconstruction of time. But calendar effects are not restricted to conventional day-of-week, month-of-year ways of deconstructing time. There can be several classes of time deconstructions, viz. based on arrangement (linear vs. cyclic) or hierarchical order of calendar. Linear time granularities respect the linear progression of time such as hours, days, weeks and months. One of the first attempts of characterizing these granularities occur in [@Bettini1998-ed]. The definitions and rules defined are inadequate to reflect periodicities in time. Hence, there is a need to define cyclic time granularities in a different approach, which can be useful in visualizing periodic behavior. Cyclic time granularities can be circular or aperiodic depending on if mappings between granularities are regular or irregular. Examples of circular can be hour of the day, day of the week and that of aperiodic granularities can be day of the month or public holidays. Time deconstructions can also be based on the hierarchical structure of time. For example, hours are nested within days, days within weeks, weeks within months, and so on. Hence, it is possible to construct single-order-up granularities like second of the minute or multiple-order-up granularities like second of the hour. lubridate [@G_Grolemund2011-vm] creates easy access and manipulation of common date-time objects. But most accessor functions are limited to single-order-up granularities.

@@

**Definition: Quasi Periodical** A granularity H is quasi-periodical with respect to a granularity G if
(1) $$G \trianglelefteq H,$$, and
(2) there exist a set of intervals E1,...,Ez (the granularity exceptions) and positive
integers R,P, where R is less than the minimum of the number of granules of H
between any 2 exceptions, such that for all $i \epsilon Z$, if $H(i) = \bigcup_{k \in [0,k]} G(j_r)$  and $H (i + R) \neq \phi$ and $i + R < min(E)$, where E is the closest existing exception after $H (i)^2$, then $H (i + R) = \bigcup_{k \in [0,k]} G(j_r + P )$.

@@

We propose a formalism of cyclic time granularities through the tsibble [@Wang2019-aw] framework of organizing temporal data. A time domain, as defined by Bettini, is essentially a mapping of the index set to the time index of a tsibble. A linear granularity is a mapping of the index set to subsets of the time domain. For example, if the time index is days, then a linear granularity might be weeks, months or years. Lining up with the the assumption in [@Bettini2000-vy] that all linear granularities can be generated from the bottom granularity by calendar algebraic operations, we assume that a bottom granularity exists and is represented by the index of the tsibble.

@@

\begin{table}[ht]
\begin{center}
\begin{tabular}{lr@{~}lr@{~}r}
\toprule
Minute-of-Hour: & $C_1(z)$ & $= z \mod 60$ & $k_1 =$&$60$ \\
Minute-of-Day: & $C_2(z)$ & $= z \mod 60*24$ & $k_2=$&$`r 60*24`$\\
Hour-of-Day: & $C_4(z)$ & $= \lfloor z/60\rfloor\mod 24$  & $k_3 =$&$24$ \\
Hour-of-Week: & $C_6(z)$ & $= \lfloor z/60\rfloor\mod 24*7$  & $k_4=$&$`r 24*7`$\\
Day-of-Week: & $C_7(z)$ & $= \lfloor z/24*60\rfloor \mod 7$ & $k_5=$&$7$\\
\bottomrule
\end{tabular}
\end{center}
\caption{Illustrative circular granularities with time index in minutes}
\label{definitions}
\end{table}

@@

Day-of-Month: & $A_1(z)$ &  $k_1=31, 30, 29, 28$\\
Hour-of-Month: & $A_2(z)$ & $k_2=24*31, 24*30, 24*29, 24*28$\\
Day-of-Year: & $A_3(z)$ & $k_3=366, 365$\\
Week-of-Month: & $A_4(z)$ & $k_4=5, 4$\\ 
\bottomrule
\end{tabular}
\end{center}
\caption{Illustrative aperiodic circular granularities with potential period lengths}
\label{aperiodic}
\end{table}

@@

\begin{equation}\label{eq3}
A_{BG, AG}(z) = BG(z - \sum_{w=0}^{k-1}\vert T_{w \mod R'} \vert) for z \in T_k
\end{equation}
where,  
z denotes the index set (index of BG),
R' denotes the number of granules of AG in each period of BG,  
$T_w$ are the sets of indices of BG describing $AG_w$ such that $AG(w) = \bigcup_{z \in T_w}BG(z)$, and $T_w = 0$ for $w \epsilon Z_{<0}$  
$\vert T_w \vert$ is the cardinality of set $T_w$.

@@

**Discussion**: Example showing aperiodic granularities relating two linear granularities each with bottom granularities are visually depicted in a series of slots in \autoref{fig:aperiodic-gran}. Each granule is represented by a box. Similar to \autoref{fig:circular-gran}, the granules that overlap share elements from the underlying time domain and the first slot shows the index set and the bottom granularity. 
Two linear granularities AG and AH are considered, where $AG = Alter^{5}_{(1,-1)}(BG, Group_3(BG))$ and $AH = Alter^{12}_{(1,-2)}(BG, Group_7(BG))$. This implies that AG is made up of shrinking every 1st granule of Group_3(BG) by 1 and AH is made up of shrinking every 1st granule of Group_3(BG) by 2. Number of granules of AG and AH in each period of BG is 2, but the number of granules of BG in each of those granules are different. A(BG, AG) and A(BG, AH) are repetitive categorisation of time, similar to circular granularities, except that the number of granules of BG is not necessarily the same across different granules of AG or AH. For AG, R' = 2, P'= 5, $T_0$ = {0,1} and $T_1$ = {2, 3, 4}. For AH, R' = 2, P'= 12, $T_0$ = {0, 1, 2, 3, 4, 5, 6} and $T_1$ = {7, 8, 9, 10, 11}, then we will have \ref{eq:eq4} and  \ref{eq:eq5}.

\begin{equation} \label{eq4}
\begin{split}
A(BG,AG)(8) & = BG(8 - \sum_{w=0}^{3-1}\vert T_{w \mod 2}\vert) , since 8 \in T_{3} \\
  & = BG(8 - \sum_{w=0}^{2}\vert T_{w\mod 2}\vert) \\
  & = BG(8 - \sum_{w=0}^{2}\vert T_{w\mod 2}\vert) \\
  & = BG(8 - \vert T_{0 \mod 2}\vert - \vert T_{1 \mod 2}\vert - \vert T_{2 \mod 2}\vert) \\
  & = BG(8 - 2*\vert T_{0 \mod 2}\vert - \vert T_{1 \mod 2}\vert) \\
  & = BG(8 - 2.2 - 3) \\
  & = BG(1) \\
\end{split}
\end{equation}

\begin{equation} \label{eq5}
\begin{split}
A(BG,AH)(10) & = BG(10 - \sum_{w=0}^{1-1}\vert T_{w \mod 2}\vert) , since 10 \in T_{1}  \\
  & = BG(10 - \sum_{w=0}^{0}\vert T_{w\mod 2}\vert) \\
  & = BG(10 - \vert T_{0 \mod 2}\vert) \\
  & = BG(10 - \vert T_{0}\vert) \\
  & = BG(10 - 7) \\
  & = BG(3) \\
\end{split}
\end{equation}

@@

If we ignore leap years and all its exception, the periodicity of the grouping (days, months), P' = 365, R' = 12, since each month would be defined as the grouping of the same number of days (31, 28, or 30, depending on the month) every year. Considering leap years and all their exceptions, period is 400 years and hence $P' = (3*365 + 366)$, $R' = 12*400$.  \autoref{eq:eq6} can be useful to find the aperiodic granularity day-of-month in both of these cases, since the grouping is periodic is both cases.

@@

The linear granularities G, H, AG and AH considered in the definitions of circular and aperiodic granularities are all periodic with respect to the bottom granularity. The only difference is G, H have regular mapping, whereas, G1 and G2 have irregular mapping with respect to BG. It can be noted here that if a linear granularity AG is quasi-periodic with respect to BG, then \ref{eq:eq6} can be modified as follows to account for exceptions $E = [e_begin, e_end]$ (\ref:{def:something})

@@

**Definition:  Order** of a linear granularity can be comprehended as the level graininess asscociated with a linear granularity. If we consider two linear granularities G and H, such that $G \preceq H$, then H is of higher order than G.

@@

Let us use the equation \ref{eq3} to compute the multiple-order-up granularity uinal_katun for Mayan calendar, which is periodic.

@@

**Example:** Considering the same example of Mayan Calendar,  it is possible to compute the single-order-up granularity tun-of-katun  given the multiple-order-up granularity uinal-baktun using equation \ref{eq5}

@@

Let us turn to Gregorian calendar for addressing this case. Suppose we have a hierarchy table for  Gregorian calendar as \ref{table_accessor2}. Since months consists of unequal number of days, any temporal unit which is higher in order than months will also have unequal number of days. This is an example of a hierarchy structure which has both circular and aperiodic single-order-up granularities. The single-order-up granularity day_month is aperiodic. Any single-order-up granularities which are formed by units below days are periodic. Similarly, all single-order-up granularities which are formed using units whose orders are higher than months are also periodic.

@@

<!-- , and indeed many parts of the world. Households are distributed geographically and have different demographic properties such as the existence of solar panels, central heating or air conditioning. The behavioral patterns in households vary substantially, for example, some famailies use a dryer for their clothes while others hang them on a line, and some households might consist of night owls, while others are morning larks.  -->

@@

Figure \ref{fig:my-hd} shows that the median is very close to the lower boundaries of all the other bands implying energy consumption for these households are left skewed. Moreover, only the pink band changes significantly in winter months (May - August). The median consumption or width of other bands doesn't vary much across seasons, implying extreme behavior or extreme customers does not vary across seasons much. Most of the behavioral changes occur in the quartiles, that too in peak hours of the day. It is to be noted here that the the level of quantiles increased too in winter months (obvious because of weather considtions), the interesting part is to notice that the relationship between bands other than quartile stayed same across seasons.

>>>>

Calendar-based graphics[@wang2018calendar] are useful in visualizing patterns in the weekly and monthly structure well and are capable of checking the weekends or special days.  Any sub-daily resolution temporal data can also be displayed using this type of faceting [@Wickham2009-pk] with days of the week, month of the year and another sub-daily deconstruction of time. But calendar effects are not restricted to conventional day-of-week, month-of-year ways of deconstructing time. There can be several classes of time deconstructions, viz. based on arrangement (linear vs. cyclic) or hierarchical order of calendar. Linear time granularities respect the linear progression of time such as hours, days, weeks and months. One of the first attempts of characterizing these granularities occur in [@Bettini1998-ed]. The definitions and rules defined are inadequate to reflect periodicities in time. Hence, there is a need to define cyclic time granularities in a different approach, which can be useful in visualizing periodic behavior. Cyclic time granularities can be circular or aperiodic depending on if mappings between granularities are regular or irregular. Examples of circular can be hour of the day, day of the week and that of aperiodic granularities can be day of the month or public holidays. Time deconstructions can also be based on the hierarchical structure of time. For example, hours are nested within days, days within weeks, weeks within months, and so on. Hence, it is possible to construct single-order-up granularities like second of the minute or multiple-order-up granularities like second of the hour. lubridate [@G_Grolemund2011-vm] creates easy access and manipulation of common date-time objects. But most accessor functions are limited to single-order-up granularities.

@@

\newtheorem{mydef}{Definition}
\begin{mydef}\label{def:quasi}
**Definition: Quasi Periodical** A granularity H is quasi-periodical with respect to a granularity G if
(1) $$G \trianglelefteq H,$$, and
(2) there exist a set of intervals E1,...,Ez (the granularity exceptions) and positive
integers R,P, where R is less than the minimum of the number of granules of H
between any 2 exceptions, such that for all $i \epsilon Z$, if $H(i) = \bigcup_{k \in [0,k]} G(j_r)$  and $H (i + R) \neq \phi$ and $i + R < min(E)$, where E is the closest existing exception after $H (i)^2$, then $H (i + R) = \bigcup_{k \in [0,k]} G(j_r + P )$.
\end{mydef}

@@

We propose a formalism of cyclic time granularities through the tsibble [@wang2019tsibble] framework of organizing temporal data. A time domain, as defined by Bettini, is essentially a mapping of the index set to the time index of a tsibble. A linear granularity is a mapping of the index set to subsets of the time domain. For example, if the time index is days, then a linear granularity might be weeks, months or years. Lining up with the the assumption in [@Bettini2000-vy] that all linear granularities can be generated from the bottom granularity by calendar algebraic operations, we assume that a bottom granularity exists and is represented by the index of the tsibble.

@@

\begin{table}[ht]
\begin{center}
\begin{tabular}{lr@{~}lr@{~}r}
\toprule
Minute-of-Hour: & $C_1(z)$ & $= z \mod 60$ & $k_1 =$&$60$ \\
Minute-of-Day: & $C_2(z)$ & $= z \mod 60*24$ & $k_2=$&$`r 60*24`$\\
Hour-of-Day: & $C_4(z)$ & $= \lfloor z/60\rfloor\mod 24$  & $k_3 =$&$24$ \\
Hour-of-Week: & $C_6(z)$ & $= \lfloor z/60\rfloor\mod 24*7$  & $k_4=$&$`r 24*7`$\\
Day-of-Week: & $C_7(z)$ & $= \lfloor z/24*60\rfloor \mod 7$ & $k_5=$&$7$\\
\bottomrule
\end{tabular}
\end{center}
\caption{Illustrative circular granularities with time index in minutes}
\label{tab:definitions}
\end{table}

@@

Day-of-Month: & $A_1(z)$ &  $k_1=31, 30, 29, 28$\\
Hour-of-Month: & $A_2(z)$ & $k_2=24*31, 24*30, 24*29, 24*28$\\
Day-of-Year: & $A_3(z)$ & $k_3=366, 365$\\
Week-of-Month: & $A_4(z)$ & $k_4=5, 4$\\ 
\bottomrule
\end{tabular}
\end{center}
\caption{Illustrative aperiodic circular granularities with potential period lengths}
\label{tab:aperiodic}
\end{table}

@@

\begin{equation}\label{eq:eq3}
A_{BG, AG}(z) = BG(z - \sum_{w=0}^{k-1}\vert T_{w \mod R'} \vert) for z \in T_k
\end{equation}
where,  
z denotes the index set (index of BG),
R' denotes the number of granules of AG in each period of BG,  
$T_w$ are the sets of indices of BG describing $AG_w$ such that $AG(w) = \bigcup_{z \in T_w}BG(z)$, and $T_w = 0$ for $w \epsilon Z_{<0}$  
$\vert T_w \vert$ is the cardinality of set $T_w$.

@@

**Discussion**: Example showing aperiodic granularities relating two linear granularities each with bottom granularities are visually depicted in a series of slots in \autoref{fig:aperiodic-gran}. Each granule is represented by a box. Similar to \autoref{fig:circular-gran}, the granules that overlap share elements from the underlying time domain and the first slot shows the index set and the bottom granularity. 
Two linear granularities AG and AH are considered, where $AG = Alter^{5}_{(1,-1)}(BG, Group_3(BG))$ and $AH = Alter^{12}_{(1,-2)}(BG, Group_7(BG))$. This implies that AG is made up of shrinking every 1st granule of Group_3(BG) by 1 and AH is made up of shrinking every 1st granule of Group_3(BG) by 2. Number of granules of AG and AH in each period of BG is 2, but the number of granules of BG in each of those granules are different. A(BG, AG) and A(BG, AH) are repetitive categorization of time, similar to circular granularities, except that the number of granules of BG is not necessarily the same across different granules of AG or AH. For AG, R' = 2, P'= 5, $T_0$ = {0,1} and $T_1$ = {2, 3, 4}. For AH, R' = 2, P'= 12, $T_0$ = {0, 1, 2, 3, 4, 5, 6} and $T_1$ = {7, 8, 9, 10, 11}, then we will have \ref{eq:eq4} and  \ref{eq:eq5}.

\begin{equation} \label{eq:eq4}
\begin{split}
A(BG,AG)(8) & = BG(8 - \sum_{w=0}^{3-1}\vert T_{w \mod 2}\vert) , 8 \in T_{3} \\
  & = BG(8 - \sum_{w=0}^{2}\vert T_{w\mod 2}\vert) \\
  & = BG(8 - \sum_{w=0}^{2}\vert T_{w\mod 2}\vert) \\
  & = BG(8 - \vert T_{0 \mod 2}\vert - \vert T_{1 \mod 2}\vert - \vert T_{2 \mod 2}\vert) \\
  & = BG(8 - 2*\vert T_{0 \mod 2}\vert - \vert T_{1 \mod 2}\vert) \\
  & = BG(8 - 2*2 - 3) \\
  & = BG(1) \\
\end{split}
\end{equation}

\begin{equation} \label{eq:eq5}
\begin{split}
A(BG,AH)(10) & = BG(10 - \sum_{w=0}^{1-1}\vert T_{w \mod 2}\vert) ,  10 \in T_{1}  \\
  & = BG(10 - \sum_{w=0}^{0}\vert T_{w\mod 2}\vert) \\
  & = BG(10 - \vert T_{0 \mod 2}\vert) \\
  & = BG(10 - \vert T_{0}\vert) \\
  & = BG(10 - 7) \\
  & = BG(3) \\
\end{split}
\end{equation}

@@

If we ignore leap years and all its exception, the periodicity of the grouping (days, months), P' = 365, R' = 12, since each month would be defined as the grouping of the same number of days (31, 28, or 30, depending on the month) every year. Considering leap years and all their exceptions, period is 400 years and hence $P' = (3*365 + 366)$, $R' = 12*400$.  \autoref{eq:eq3} can be useful to find the aperiodic granularity day-of-month in both of these cases, since the grouping is periodic is both cases.

@@

The linear granularities G, H, AG and AH considered in the definitions of circular and aperiodic granularities are all periodic with respect to the bottom granularity. The only difference is G, H have regular mapping, whereas, G1 and G2 have irregular mapping with respect to BG. It can be noted here that if a linear granularity AG is quasi-periodic with respect to BG, then \ref{eq:eq3} can be modified as follows to account for exceptions $E = [e_{begin}, e_{end}]$ (\autoref{def:quasi})

@@

**Definition:  Order** of a linear granularity can be comprehended as the level graininess associated with a linear granularity. If we consider two linear granularities G and H, such that $G \preceq H$, then H is of higher order than G.

@@

Let us use the equation \ref{eq:eq3} to compute the multiple-order-up granularity uinal_katun for Mayan calendar, which is periodic.

@@

**Example:** Considering the same example of Mayan Calendar,  it is possible to compute the single-order-up granularity tun-of-katun  given the multiple-order-up granularity uinal-baktun using equation \ref{eq:eq5}

@@

Let us turn to Gregorian calendar for addressing this case. Suppose we have a hierarchy table using some linear granularties from Gregorian calendar. Since months consists of unequal number of days, any temporal unit which is higher in order than months will also have unequal number of days. This is an example of a hierarchy structure which has both circular and aperiodic single-order-up granularities. The single-order-up granularity day_month is aperiodic. Any single-order-up granularities which are formed by units below days are periodic. Similarly, all single-order-up granularities which are formed using units whose orders are higher than months are also periodic.

@@

<!-- , and indeed many parts of the world. Households are distributed geographically and have different demographic properties such as the existence of solar panels, central heating or air conditioning. The behavioral patterns in households vary substantially, for example, some families use a dryer for their clothes while others hang them on a line, and some households might consist of night owls, while others are morning larks.  -->

@@

Figure \ref{fig:my-hd} shows that the median is very close to the lower boundaries of all the other bands implying energy consumption for these households are left skewed. Moreover, only the pink band changes significantly in winter months (May - August). The median consumption or width of other bands doesn't vary much across seasons, implying extreme behavior or extreme customers does not vary across seasons much. Most of the behavioral changes occur in the quartiles, that too in peak hours of the day. It is to be noted here that the the level of quantiles increased too in winter months (obvious because of weather conditions), the interesting part is to notice that the relationship between bands other than quartile stayed same across seasons.
